To be FAIR: Theory Specification Needs an Update

### 2025-02-25 manuscript.Rmd
Update manuscript.Rmd

In a progressive research program [@lakatosHistoryScienceIts1971],
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
<!-- NvD: This is not exactly what Lakatos said. Also, I don't think many psychologists are familiar with his work. We do not need to reference his perspective. -->
<!-- by successively increasing the empirical content of a theory. -->
<!-- In psychology, this is typically achieved by either generalizing to a larger set of people, contexts, situations, or environments. <!--AB would you agree?-->
<!-- CJ: I am not sure if I agree; I think increasing the empirical content rather implies more and more precision, which loses generality but gains specificity. Not sure how to resolve.
This needs more work! -->
<!-- NvD: SOmething like? "In the envisioned ideal, the empirical cycle is a virtuous spiral where theories become increasingly more precise and/or universal (i.e., Popper's empirical content), offering hypotheses that can be more easily falsified / severely tested, data become more informative / less ambiguous about the patterns in nature, which in turn inform even more precise and/or universal theories. -->
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature: 
According to @kuhbergerPublicationBiasPsychology2014, 89.6% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis [@mcphetresDecadeTheoryReflected2021].
The remaining 85% of deductive studies lacked an explicit connection between theory and hypothesis.
Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
in which case it would be straightforward and important to make these explicit [@friedTheoriesModelsWhat2020, @norouziCapturingCausalClaims2024].
Or, perhaps some hypotheses are thoughtlessly reported as part of entrenched research practices [@gigerenzerNullRitualWhat2004],
but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected [@vanlissaTeacherCornerEvaluating2020].
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
<!-- NvD: I don't think that statement completely covers it. How about: "Testing ad-hoc hypotheses not grounded in theory might inform us of the psychological phenomena that we can consistently observe (e.g., cognitive biases, attentional blindess, comorbidity of depression and anxiety disorder, etc.). However, it does not advance our understanding of the processes and mechanisms that bring about these phenomena. -->
Put differently: collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings [@vanrooijTheoryTestHow2021].

@@

The present paper introduces open science methods for theory specification and archival.
Merely publishing theory in a research article does not make it open;
to be open, theory should adhere to established open science standards.
We do so by implementing theories as information artifacts,
archived with apropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
Metadata are "data about the data", they provide information about the nature and content of an information artifact.
Metadata are stored in the repository where the version of record of the FAIR theory is deposited.
FAIR theories are *Findable* via a DOI or by searching the repository they are archived in;
*Accessible* in a machine- and human-readable filetype;
*Interoperable* for specific purposed, for example, within the data analysis environment;
and *Reusable* in the practical and legal sense, so that they may be iteratively improved by the author or by others.
<!-- AB I think some initial word on what is covered is needed here; otherwise readers may wonder what we mean to cover--> 
<!-- CJ: Since this is in the next paragraph, I think it's redundant. Will merge there -->
Following the original proposal of Lamprecht and colleagues,
we adapt the FAIR principles for theory, see [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv).
We reflect on the necessary changes (which are minor),
as well as on the current state and future of FAIR theory in psychology.
The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
and we provide worked examples to encourage their adoption.

@@

<!-- We don't explicitly define theory, but discuss the most important ideas of what theory might be, and advocate for a pluralism of information that can be tracked as FAIR theory (e.g. also such things as scientific practice (experimental design, materials)). -->
<!-- The nature of the tracked information can change over time, as a scientific theory matures. -->
<!-- The discussion about the distinction between FAIR theories and models is removed, as I believe it might be confusing for first-time readers and not really necessary for the bigger picture of the paper (but of course could be re-added). -->
<!-- NvD: I wonder if we need what is stated about the syntactic, semantic, and pragmatic perspectives. We can leave it at the characterization that theories are (typically) concerned with the explanation and understanding of what can be observed (directly or indirectly) tested (in the laboratory or natural experiments) and manipulated. But, this can come in many forms and we do not exclude the possibility that there are acceptions to this characterization that we could still consider a scientific theory. We can then state that we use Meehl's view for convenience in this paper, but that it is compatible other perspectives (or something.  -->
Definitions of theory abound and are the subject of extensive scholarly debate.
Given that a pluriformity of definitions are consistent with FAIR theory principles,
the method is not contained to any one particular definition.
Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories]
The syntactic view describes theories as "sets of sentences in a given logical domain language" [@scientifictheories, ch. 2],
acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
We recognize the syntactic view in Meehl's hierarchy of ever-more specific "statements" a theory might contain [-@meehlAppraisingAmendingTheories1990]:
statements about the types of entities postulated (i.e., ontology),
statements about causal connections between those entities,
statements about the functional form of those connections,
and statements about their specific numerical values [cf. @frankenhuisStrategicAmbiguitySocial2023; @guestWhatMakesGood2024].

The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
It shifts the focus from theories as collections of sentences to mathematical models.
The term "model" is not uniquely defined within the literature;
it has been described as a "specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [@friedTheoriesModelsWhat2020].
This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
<!-- Paradigms at the broadest level encompass the larger narrative, frameworks gather and organize concepts and terms, but only theories have a degree of specificity that allows us to derive specific testable and falsifiable hypotheses or models with exact mathematical relations between concepts (cf. Partelow, 2023, Table 1, https://link.springer.com/article/10.1007/s13412-023-00833-w#). -->
The pragmatic view holds that there might not be one structure or definition of scientific theories, but instead, definitions differ across scientific domains.
It also argues that nonformal aspects (e.g. commonly used analogies) and practices (e.g. experimental designs) can be an important part of scientific theories.

It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be represented as FAIR theory. 
As the practice of FAIRification becomes more embedded,
we expect that it will become increasingly clear what kind and form of information is useful.
As a particular FAIR theory evolves, details may be added, and the nature of the information tracked might even change.
For example, following Meehl, we could envision a theory that starts out with establishing, through observation, an ontology of constructs relevant for a given phenomenon.
After initial exploratory research, the theory might be further specified by making assumptions about how these constructs are causally connected.
Over time, more precise *statistical/mathematical models* could be derived by further assuming a specific functional form for relationships (e.g., linear effects) and error families for the distribution of measured variables (e.g., normal distributions).
This allows for the specification of statistical models, which make just enough assumptions to allow the estimation of the remaining unknown parameters (e.g., regression slopes) from data.
Going even further, a *generative/computational model* could be specified,
which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
Also, aspects of scientific practice might be added over time - either to the theory itself, or as references recorded in the theory metadata.
Examples include experimental designs (e.g. longitudinal designs observing change over time), measurement tools (e.g. different questionnaires used to assess the same construct), or study subjects (e.g. specific strains of rats).

@@

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use. 
<!-- NvD: Maybe we should add "specification" as part of "formalization" or "formalisation" as a part of "specification", because not everybody that is advocating better theories thinks that we should focus on formalization.  -->
Greater formalization increases theories' *empirical content* [@popperLogicScientificDiscovery2002] because it expresses ideas as precise statements, clearly demarcating what should (not) be observed.
For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the specification of the decay rate, recall success, or rehearsal sequence [@lewandowsky2010computational].
Without committing to specific implementations a-priori,
the theory becomes hard to test.
Compared to theories expressed in natural language,
formal theories facilitate inconsistency checking and evaluation of a theory's (lack of) ambiguity.
Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
More precise theories are easier to falsify,
which necessitates specific revisions and advances our principled understanding of the phenomena they describe.

FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
rather, it increases the fidelity and ease with which they are communicated.
Thus, FAIR theory does not require theories to be formal,
and formal theory can be represented in a way that is not FAIR.
The FAIR principles pertain to theories' metadata documentation and sharing in digital archives, with the aim of enhancing their reusability and extensibility.
They can be applied to theories representated in natural language,
as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
Thus, for example, "grounded theory", derived from qualitative research,
can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
Conversely, a formal theory is not FAIR if it is confined to in a journal article without any key words to identify it as a theory paper (lacking Findability), represented as a bitmap image (limiting Accessibility and Interoperability), and behind a paywall (limiting Reusability).
FAIR theory is thus consistent with, but does not require, formalization (also see *Accessibility*).

>>>>

In a progressive research program [@lakatosHistoryScienceIts1971],
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
<!-- NvD: This is not exactly what Lakatos said. Also, I don't think many psychologists are familiar with his work. We do not need to reference his perspective. -->
<!-- by successively increasing the empirical content of a theory. -->
<!-- In psychology, this is typically achieved by either generalizing to a larger set of people, contexts, situations, or environments. <!--AB would you agree?-->
<!-- CJ: I am not sure if I agree; I think increasing the empirical content rather implies more and more precision, which loses generality but gains specificity. Not sure how to resolve.
This needs more work! -->
<!-- NvD: SOmething like? "In the envisioned ideal, the empirical cycle is a virtuous spiral where theories become increasingly more precise and/or universal (i.e., Popper's empirical content), offering hypotheses that can be more easily falsified / severely tested, data become more informative / less ambiguous about the patterns in nature, which in turn inform even more precise and/or universal theories. -->
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature: 
According to @kuhbergerPublicationBiasPsychology2014, 89.6% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis [@mcphetresDecadeTheoryReflected2021].
The remaining 85% of deductive studies lacked an explicit connection between theory and hypothesis.
Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
in which case it would be straightforward and important to make these explicit [@friedTheoriesModelsWhat2020; @norouziCapturingCausalClaims2024].
Or, perhaps some hypotheses are thoughtlessly reported as part of entrenched research practices [@gigerenzerNullRitualWhat2004],
but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected [@vanlissaTeacherCornerEvaluating2020].
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
<!-- NvD: I don't think that statement completely covers it. How about: "Testing ad-hoc hypotheses not grounded in theory might inform us of the psychological phenomena that we can consistently observe (e.g., cognitive biases, attentional blindess, comorbidity of depression and anxiety disorder, etc.). However, it does not advance our understanding of the processes and mechanisms that bring about these phenomena. -->
Put differently: collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings [@vanrooijTheoryTestHow2021].

@@

The present paper introduces open science methods for theory specification and archival.
Merely publishing theory in a research article does not make it open;
to be open, theory should adhere to established open science standards.
We do so by implementing theories as information artifacts,
archived with apropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
Metadata are "data about the data", they provide information about the nature and content of an information artifact.
Metadata are stored in the repository where the version of record of the FAIR theory is deposited.
FAIR theories are *Findable* via a DOI or by searching the repository they are archived in;
*Accessible* in a machine- and human-readable filetype;
*Interoperable* for specific purposes, for example, within the data analysis environment;
and *Reusable* in the practical and legal sense, so that they may be iteratively improved by the author or by others.
<!-- AB I think some initial word on what is covered is needed here; otherwise readers may wonder what we mean to cover--> 
<!-- CJ: Since this is in the next paragraph, I think it's redundant. Will merge there -->
Following the original proposal of Lamprecht and colleagues,
we adapt the FAIR principles for theory, see [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv).
We reflect on the necessary changes (which are minor),
as well as on the current state and future of FAIR theory in psychology.
The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
and we provide worked examples to encourage their adoption.

@@

<!-- We don't explicitly define theory, but discuss the most important ideas of what theory might be, and advocate for a pluralism of information that can be tracked as FAIR theory (e.g. also such things as scientific practice (experimental design, materials)). -->
<!-- The nature of the tracked information can change over time, as a scientific theory matures. -->
<!-- The discussion about the distinction between FAIR theories and models is removed, as I believe it might be confusing for first-time readers and not really necessary for the bigger picture of the paper (but of course could be re-added). -->
<!-- NvD: I wonder if we need what is stated about the syntactic, semantic, and pragmatic perspectives. We can leave it at the characterization that theories are (typically) concerned with the explanation and understanding of what can be observed (directly or indirectly) tested (in the laboratory or natural experiments) and manipulated. But, this can come in many forms and we do not exclude the possibility that there are acceptions to this characterization that we could still consider a scientific theory. We can then state that we use Meehl's view for convenience in this paper, but that it is compatible other perspectives (or something.  -->
Definitions of theory are abundant, and are the subject of extensive scholarly debate.
Given that a pluriformity of definitions are consistent with FAIR theory principles,
our suggested approach is not limited to any one particular definition.
Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories].
The syntactic view describes theories as "sets of sentences in a given logical domain language" [@scientifictheories, ch. 2],
acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
We recognize the syntactic view in Meehl's [-@meehlAppraisingAmendingTheories1990] hierarchy of ever-more specific "statements" a theory might contain:
statements about the types of entities postulated (i.e., ontology),
statements about causal connections between those entities,
statements about the functional form of those connections,
and statements about their specific numerical values [cf. @frankenhuisStrategicAmbiguitySocial2023; @guestWhatMakesGood2024].
The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
It shifts the focus from theories as collections of sentences to mathematical models.
The term "model" is not uniquely defined within the literature;
it has been described as a "specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [@friedTheoriesModelsWhat2020].
This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
<!-- Paradigms at the broadest level encompass the larger narrative, frameworks gather and organize concepts and terms, but only theories have a degree of specificity that allows us to derive specific testable and falsifiable hypotheses or models with exact mathematical relations between concepts (cf. Partelow, 2023, Table 1, https://link.springer.com/article/10.1007/s13412-023-00833-w#). -->
The pragmatic view holds that there might not be one structure or definition of scientific theories, but instead, definitions differ across scientific domains.
It also argues that nonformal aspects (e.g. commonly used analogies) and practices (e.g. experimental designs) can be an important part of scientific theories.

It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be represented as FAIR theory. 
As the practice of FAIRification becomes more embedded,
we expect that it will become increasingly clear what kind and form of information is useful.
As a particular FAIR theory evolves, details may be added, and the nature of the information tracked might even change.
For example, following Meehl, we could envision a theory that starts out with establishing, through observation, an ontology of constructs relevant for a given phenomenon.
After initial exploratory research, the theory might be further specified by making assumptions about how these constructs are causally connected.
Over time, more precise *statistical/mathematical models* could be derived by further assuming a specific functional form for relationships (e.g., linear effects) and error families for the distribution of measured variables (e.g., normal distributions).
This allows for the specification of statistical models, which make just enough assumptions to allow the estimation of the remaining unknown parameters (e.g., regression slopes) from data.
Going even further, a *generative/computational model* could be specified,
which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
Also, aspects of scientific practice might be added over time - either to the theory itself, or as references recorded in the theory metadata.
Examples include experimental designs (e.g., longitudinal designs observing change over time), measurement tools (e.g., different questionnaires used to assess the same construct), or study subjects (e.g., specific strains of rats).

@@

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use. 
<!-- NvD: Maybe we should add "specification" as part of "formalization" or "formalisation" as a part of "specification", because not everybody that is advocating better theories thinks that we should focus on formalization.  -->
More rigorous formalization increases theories' *empirical content* [@popperLogicScientificDiscovery2002] because it expresses ideas as precise statements, clearly demarcating what should be observed and what not if the theory were true.
For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for its particular clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory [@lewandowsky2010computational].
Without committing to specific implementations a-priori,
the theory becomes hard to test.
Compared to theories expressed in natural language,
formal theories facilitate inconsistency checking and evaluation of a theory's (lack of) ambiguity.
Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
More precise theories are easier to falsify,
which necessitates specific revisions and advances our principled understanding of the phenomena they describe.

FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
rather, it increases the fidelity and ease with which they are communicated.
Thus, FAIR theory does not require theories to be formal.
At the same time, not every formal theory is FAIR.
The FAIR principles pertain to theories' metadata documentation and sharing in digital archives, with the aim of enhancing their reusability and extensibility.
They can be applied to theories representated in natural language,
as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
Thus, for example, "grounded theory", derived from qualitative research,
can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
Conversely, a formal theory is not FAIR if it is confined to in a journal article without any key words to identify it as a theory paper (lacking Findability), represented as a bitmap image (limiting Accessibility and Interoperability), and behind a paywall (limiting Reusability).
FAIR theory is thus consistent with, but does not require, formalization (also see *Accessibility*).
